The Proofs

### 2024-07-31 proofs/content.tex
prettify + minor edits

\begin{algorithm}[Type Well-formedness]
  \label{alg:wf}
  \hfill

  \begin{multicols}{2}
  \ottdefnWFTNiWFLabeled{}
  \columnbreak

  \ottdefnWFTPiWFLabeled{}
  \end{multicols}

@@

Sometimes it is convenient to construct substitution 
explicitly mapping each variable from a list (or a set)
to a type. Such substitutions are denoted as $[[iPs / pas]]$
and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
the corresponding types.
\begin{definition}[Explicit Substitution]
  \hfill
  \begin{itemize}
    \item [$-$]
      Suppose that $[[nas]]$ is a list of negative type variables,
      and $[[iNs]]$ is a list of negative types of the same length.
      Then $[[iNs / nas]]$ denotes a substitution such that 
      \begin{enumerate}
        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
      \end{enumerate}
    \item [$+$]
      Positive explicit substitution $[[iPs / pas]]$
      is defined symmetrically.
  \end{itemize}
\end{definition}

@@

\begin{definition} 
  \label{def:subDOne}
  \hfill

  \begin{multicols}{2}
    \ottdefnDOneNsubLabeled{}

    \ottdefnDOnePsupLabeled{}
  \end{multicols}
  \hfill

  \begin{multicols}{2}
    \ottdefnDOneNeqLabeled{}

    \ottdefnDOnePeqLabeled{}
  \end{multicols}
\end{definition}

@@

\begin{definition}[Algorithmic Types]
  \label{def:algo-types}
  \hfill\\
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottuN\ottinterrule
    }

    \ottgrammartabular{
      \ottuP\ottinterrule
    }
    \columnbreak
  \end{multicols}
\end{definition}

@@

  \begin{multicols}{2}
  \ottdefnWFATNauWFLabeled{}
  \columnbreak

  \ottdefnWFATPauWFLabeled{}
  \end{multicols}

Algorithmic Type Context are used in the unification algorithm.
In the subtyping algorithm, 
the context needs to remember additional information.
In the subtyping context, each algorithmic variable is associated with a
context it must be instantiated in 
(i.e. the context in which the type replacing the variable must be well-formed).
This association is represented by \emph{algorithmic subtyping context} $[[Θ]]$.
\begin{definition}[Algorithmic Subtyping Context $[[Θ]]$]
  \hfill \\
  Algorithmic Subtyping Context $[[Θ]]$ is represented by a set of 
  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
  and $[[Γ]]$ is a context in which they must be instantiated.
  We assume that no two entries associating the same variable
  appear in $[[Θ]]$.

@@

  \ottdefnONVarLabeled{}
  \ottdefnOPVarLabeled{}

@@

  \begin{multicols}{2}
  \ottdefnNrmNNormLabeled{}
  \columnbreak\\
  \ottdefnNrmPNormLabeled{}
  \end{multicols}

  Analogously, we define the normalization of algorithmic types by adding base cases:

  \begin{multicols}{2}
  \ottdefnNrmuNNormLabeled{}
  \columnbreak\\
  \ottdefnNrmuPNormLabeled{}
  \end{multicols}

@@

The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
the subtyping context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
and $[[uP]]$ and $[[iQ]]$ for the positive case. 
As one of the invariants, we require
$[[iM]]$ and $[[iQ]]$ to be declarative (i.e. not containing algorithmic variables).
The output of the algorithm is a set of \emph{subtyping constraints} $[[SC]]$,
which will be discussed in the next section.

@@

  \item \ruleref{\ottdruleAArrowLabel}.
    In this case, the algorithm makes two calls:
    a recursive call to the negative subtyping algorithm for the argument types,
    and a call to the positive subtyping algorithm for the result types.
    After that, the resulting constraints are merged using the
    \emph{subtyping constraint merge} procedure, 
    which is discussed in \cref{sec:a:constraint-merge}.
  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
    are symmetric. These are the only places where 
    the algorithmic variables are introduced.
    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
    of the quantified variables: these variables are replaced by 
    fresh algorithmic variables in the body of the quantified type,
    the algorithmic variables are added to the subtyping context $[[Θ]]$,
    after that, the recursive call is made. Notice that the declarative context
    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
    which matches the declarative system.
\end{itemize}

@@

\begin{definition}[Well-formed Constraint]
  We say that a constraint is well-formed in a
  subtyping context $[[Θ]]$ if all its entries are well-formed in
  the corresponding elements of $[[Θ]]$.
  More formally, 
  $[[Θ ⊢ SC]]$ holds iff for every $[[scE]] \in $ $[[SC]]$,
  such that $[[scE]]$ restricts $[[α̂±]]$,
  we have $[[Θ(α̂±) ⊢ scE]]$.

@@

\begin{figure}
  \begin{minipage}[t]{0.39\linewidth}
    \vspace*{0pt}
    \ottgrammartabular{
      \ottfT\ottprodnewline
      \ottprodline{}{}{}{}{}{}\ottprodnewline
      \ottprodline{}{}{}{}{}{}\ottprodnewline
      \ottprodline{|}{[[∃as.fT]]}{}{}{}{}\ottprodnewline
      \ottprodline{}{\equiv [[∀b.(∀as.(fT → b)) → b]]}{}{}{}{}\ottinterrule
    }
  \end{minipage}
  \begin{minipage}[t]{0.6\linewidth}
    \vspace*{0pt}
    \ottgrammartabular{
      \ottft\ottprodnewline
      \ottprodline{}{}{}{}{}{}\ottprodnewline
      \ottprodline{|}{[[ft' ● fargs]]}{\equiv (([[ft']] [[ft1]])\, \dots )\,[[ftn]]}{}{}{}\ottprodnewline
      \ottprodline{|}{[[let x = ft1 ; ft2]]} {\equiv [[(λx.ft2) ft1]]}{}{}{}\ottprodnewline
      \ottprodline{|}{[[pack ft : fT]]}{\equiv [[Λb. λf. f ft]]}{}{}{}\ottprodnewline
      \ottprodline{|}{[[unpack ( as , x ) = ft1 ; ft2]]}{\equiv [[ft1 (Λ as. λx. ft2)]]}{}{}{}\ottinterrule
    }
  \end{minipage}
  \caption{Grammar of \systemf}
  \label{fig:systemf-grammar}
\end{figure}

@@

\begin{observation}
  The following rules are admissible in \systemf: \hfill

  \begin{minipage}{0.5\textwidth}
    \ottusedrule{\ottdruleSFtyAUnpackLabeled{}}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \ottusedrule{\ottdruleSFtyAPackLabeled{}}
    \ottusedrule{\ottdruleSFtyALetLabeled{}}
  \end{minipage}
\end{observation}

@@

\begin{figure}[t]
  \begin{minipage}[t]{0.65\linewidth}
    \centering
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization}
    \label{fig:depolarization}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[t]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}
\end{figure}

>>>>

\pagebreak[4]
\begin{algorithm}[Type Well-formedness]
  \label{alg:wf}
  \hfill\\
  \begin{minipage}[t]{0.45\textwidth}
    \ottdefnWFTNiWFLabeled{}
  \end{minipage}
  \begin{minipage}[t]{0.45\textwidth}
    \ottdefnWFTPiWFLabeled{}
  \end{minipage}
\end{algorithm}

@@

Sometimes it is convenient to construct a substitution 
that explicitly maps each variable \emph{from a list} 
(or a set) to a type.
Such substitutions are denoted as $[[iPs / pas]]$
and $[[iNs / nas]]$, where $[[iPs]]$ and $[[iNs]]$ are lists of 
the corresponding types.
\begin{definition}[Explicit Substitution]
  \hfill
  \begin{itemize}
    \item [$-$]
      Suppose that $[[nas]]$ is a list of negative type variables,
      and $[[iNs]]$ is a list of negative types of the same length.
      Then $[[iNs / nas]]$ denotes a substitution such that 
      \begin{enumerate}
        \item for $[[αi⁺ ∊ {nas}]]$, $[[ [iNs / nas] αi⁺]] = [[iNi]]$;
        \item for $[[β⁺ ∉ {nas}]]$, $[[ [iNs / nas] β⁺]] = [[β⁺]]$.
      \end{enumerate}
    \item [$+$]
      Positive explicit substitution $[[iPs / pas]]$
      is defined symmetrically.
  \end{itemize}
\end{definition}

@@

\begin{definition}[Declarative Subtyping and Equivalence]
  \label{def:subDOne}
  \hfill\\
  \begin{minipage}[t]{0.45\textwidth}
      \ottdefnDOneNsubLabeled{}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.48\textwidth}
      \ottdefnDOnePsupLabeled{}
  \end{minipage}

  \begin{minipage}[t]{0.45\textwidth}
      \ottdefnDOneNeqLabeled{}
  \end{minipage}
  \hfill
  \begin{minipage}[t]{0.45\textwidth}
      \ottdefnDOnePeqLabeled{}
  \end{minipage}
\end{definition}

@@

\begin{definition}[Algorithmic Types]
  \label{def:algo-types}
  \hfill
  \begin{multicols}{2}
    \ottgrammartabular{
      \ottuN\ottinterrule
    }
    \columnbreak\\
    \ottgrammartabular{
      \ottuP\ottinterrule
    }
  \end{multicols}
  \hfill\\
\end{definition}

@@

  \begin{minipage}[t]{0.48\textwidth}
    \ottdefnWFATNauWFLabeled{}
  \end{minipage}%
  \hfill
  \begin{minipage}[t]{0.48\textwidth}
    \ottdefnWFATPauWFLabeled{}
  \end{minipage}
\end{algorithm}

Algorithmic Type Contexts are used in the unification algorithm.
In the subtyping algorithm, 
the context needs to remember additional information.
In particular, each algorithmic variable is associated with a
context it must be instantiated in 
(i.e. the context in which the type replacing the variable must be well-formed).
This association is represented by \emph{instantiation context} $[[Θ]]$.
\begin{definition}[Instantiation context $[[Θ]]$]
  \hfill \\
  Instantiation context $[[Θ]]$ is represented by a set of 
  entries of form $[[ α̂⁺[Γ] ]]$ and $[[ α̂⁻[Γ] ]]$,
  where $[[α̂⁺]]$ and $[[α̂⁻]]$ are algorithmic variables,
  and $[[Γ]]$ is a context in which they must be instantiated.
  We assume that $[[Θ]]$ does not contain multiple entries for the same variable.

@@

  \ottdefnONVarLabeled{}
  \newpage
  \ottdefnOPVarLabeled{}

@@

  \begin{minipage}[t]{0.48\textwidth}
    \ottdefnNrmNNormLabeled{}
  \end{minipage}%
  \hfill
  \begin{minipage}[t]{0.48\textwidth}
    \ottdefnNrmPNormLabeled{}
  \end{minipage}

  Analogously, we define the normalization of algorithmic types by adding the following base cases:

  \begin{minipage}[t]{0.48\textwidth}
    \ottdefnNrmuNNormLabeled{}
  \end{minipage}%
  \hfill
  \begin{minipage}[t]{0.48\textwidth}
    \ottdefnNrmuPNormLabeled{}
  \end{minipage}

@@

The inputs of the subtyping algorithm are the declarative context $[[Γ]]$,
the instantiation context $[[Θ]]$ (it specifies in which contexts the algorithmic variables
must be instantiated), and the types themselves: $[[uN]]$ and $[[iM]]$ for the negative case,
and $[[uP]]$ and $[[iQ]]$ for the positive case. 
As one of the invariants, we require
$[[iM]]$ and $[[iQ]]$ to be declarative (i.e. not containing algorithmic variables).
The output of the algorithm is a set of \emph{subtyping constraints} $[[SC]]$,
which will be discussed in the next section.

@@

  \item \ruleref{\ottdruleAArrowLabel}.
    In this case, the algorithm makes two calls:
    a recursive call to the negative subtyping algorithm for the argument types,
    and a call to the positive subtyping algorithm for the result types.
    After that, the resulting constraints are merged using the
    \emph{subtyping constraint merge} procedure, 
    which is discussed in \cref{sec:a:constraint-merge}.
  \item \ruleref{\ottdruleAForallLabel} and \ruleref{\ottdruleAExistsLabel}
    are symmetric. These are the only places where 
    the algorithmic variables are introduced.
    It is done by algorithmization (\cref{sec:variable-algorithmization}) 
    of the quantified variables: these variables are replaced by 
    fresh algorithmic variables in the body of the quantified type,
    the algorithmic variables are added to the instantiation context $[[Θ]]$,
    after that, the recursive call is made. Notice that the declarative context
    $[[Γ]]$ is extended by the quantified variables from the right-hand side,
    which matches the declarative system.
\end{itemize}

@@

\begin{definition}[Well-formed Constraint]
  We say that a constraint is well-formed in an 
  instantiation context $[[Θ]]$ if all its entries are well-formed in
  the corresponding elements of $[[Θ]]$.
  More formally, 
  $[[Θ ⊢ SC]]$ holds iff for every $[[scE]] \in $ $[[SC]]$,
  such that $[[scE]]$ restricts $[[α̂±]]$,
  we have $[[Θ(α̂±) ⊢ scE]]$.

@@

\newcommand{\graytxt}[1]{\textcolor{darkgray}{#1}}
\newcommand{\ottprodlinecomm}[7]{\graytxt{#1}& & \graytxt{$#2$} & \graytxt{$#3$} & \graytxt{$#4 #5$} & \graytxt{$#6$} & \graytxt{$#7$}}

\begin{figure}
    \ottgrammartabular{
      \ottfT\ottprodnewline
      \\
      \ottprodlinecomm{(syntax sugar)}{|}{[[∃as.fT]]}{\equiv [[∀b.(∀as.(fT → b)) → b]]}{}{}{}\ottinterrule
    }

    \ottgrammartabular{
      \ottft\ottprodnewline
      \\
      \ottprodlinecomm{(syntax sugar)}{|}{[[ft' ● fargs]]}{\equiv (([[ft']] [[ft1]])\, \dots )\,[[ftn]]}{}{}{}\ottprodnewline
      \ottprodlinecomm{}{|}{[[let x = ft1 ; ft2]]} {\equiv [[(λx.ft2) ft1]]}{}{}{}\ottprodnewline
      \ottprodlinecomm{}{|}{[[pack ft : fT]]}{\equiv [[Λb. λf. f ft]]}{}{}{}\ottprodnewline
      \ottprodlinecomm{}{|}{[[unpack ( as , x ) = ft1 ; ft2]]}{\equiv [[ft1 (Λ as. λx. ft2)]]}{}{}{}\ottinterrule
    }

  \caption{Grammar of \systemf}
  \label{fig:systemf-grammar}
\end{figure}

@@

\newpage
\begin{observation}
  The following rules are admissible in \systemf: \hfill\\
  \begin{minipage}{0.5\textwidth}
    \ottusedrule{\ottdruleSFtyAUnpackLabeled{}}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \ottusedrule{\ottdruleSFtyAPackLabeled{}}
    \ottusedrule{\ottdruleSFtyALetLabeled{}}
  \end{minipage}
\end{observation}

@@

\begin{figure}[t]
  \begin{minipage}[b]{0.65\linewidth}
    \centering
      \begin{minipage}[t]{0.4\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \begin{minipage}[t]{0.4\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization}
    \label{fig:depolarization}
  \end{minipage}
  \begin{minipage}[b]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}
\end{figure}

### 2024-07-31 proofs/tex/alg-neg-subtyping-lemmas.tex
prettify + minor edits

            This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
            obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2 : uv(uN')]]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
            Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1 : uv(uP)]]$
            and $[[ Θ ⊢ uσ : SC1 ]]$.

            Finally, we need to show that the merge of $[[SC1]]$ and $[[SC2]]$ is successful and
            satisfies the required properties.
            To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness})
            (notice that $[[Θ ⊢ uσ : uv(uP → uN')]]$ means 
            $[[Θ ⊢ uσ : uv(uP) ∪ uv(uN')]]$).
            This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined and $[[ Θ ⊢ uσ : SC ]]$ holds.

>>>>

            This way, we can apply the induction hypothesis to $[[Γ ⊢ [uσ]uN' ≤ iM']]$ to 
            obtain $[[Γ ; Θ ⊨ uN' ≤ iM' ⫤ SC2]]$ such that $[[Θ ⊢ SC2 : uv uN']]$ and $[[ Θ ⊢ uσ : SC2 ]]$.
            Also, we can apply the completeness of the positive subtyping (\cref{lemma:pos-subt-completeness}) to 
            $[[ Γ ⊢ [uσ]uP ≥ iQ ]]$ to obtain $[[Γ ; Θ ⊨ uP ≥ iQ ⫤ SC1]]$ such that $[[Θ ⊢ SC1 : uv uP]]$
            and $[[ Θ ⊢ uσ : SC1 ]]$.

            Finally, we need to show that the merge of $[[SC1]]$ and $[[SC2]]$ is successful and
            satisfies the required properties.
            To do so, we apply the completeness of subtyping constraint merge (\cref{lemma:merge-completeness})
            (notice that $[[Θ ⊢ uσ : uv uP → uN']]$ means 
            $[[Θ ⊢ uσ : uv uP ∪ uv uN']]$).
            This way, $[[Θ ⊢ SC1 & SC2 = SC]]$ is defined and $[[ Θ ⊢ uσ : SC ]]$ holds.

### 2024-07-31 proofs/tex/alg-neg-subtyping-statements.tex
prettify + minor edits

\begin{restatable}[Soundness of Negative Subtyping]{lemma}{lemNegSubtSoundness}
    \label{lemma:neg-subt-soundness}
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
    $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
    $[[Θ ⊢ SC : uv(uN)]]$ and 
    for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{restatable}

\begin{restatable}[Completeness of the Negative Subtyping]{lemma}{lemNegSubtCompleteness}
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv(uN)]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{restatable}

>>>>

\begin{restatable}[Soundness of Negative Subtyping]{lemma}{lemNegSubtSoundness}
    \label{lemma:neg-subt-soundness}
    If $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$ and 
    $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$, then 
    $[[Θ ⊢ SC : uv uN]]$ and 
    for any normalized $[[uσ]]$ such that $[[ Θ  ⊢ uσ : SC ]]$,
    $[[ Γ ⊢ [uσ]uN ≤ iM ]]$.
\end{restatable}

\begin{restatable}[Completeness of the Negative Subtyping]{lemma}{lemNegSubtCompleteness}
    \label{lemma:neg-subt-completeness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Γ ⊢ iM]]$, $[[Γ ; dom(Θ) ⊢ uN]]$,
    and $[[uN]]$ does not contain negative unification variables ($[[α̂⁻]] \notin [[uv uN]]$).
    Then for any $[[Θ ⊢ uσ : uv uN]]$ such that $[[Γ ⊢ [uσ]uN ≤ iM]]$,
    there exists $[[Γ ; Θ ⊨ uN ≤ iM ⫤ SC]]$ and moreover, $[[ Θ ⊢ uσ : SC ]]$.
\end{restatable}
